---
title: "Effectiveness of Contraception Methods"
author: Travis Dawry
date: May 8, 2022
format: pdf
editor: visual
---

## Purpose

Visualization to show the expected failure of typical contraception methods.

Percentage of women experiencing an unintended pregnancy during the first year of typical use.

Probability of at least one failure over N years, with typical use.

## Source

Source of data is *Contraceptive failure in the United States*. Trussell,J. Contraception,2011,Vol.83(5),p.398

## Code

```{r libraries}
library(forcats)
library(ggplot2)
library(viridis)
library(tibble)
library(tidyr)
```

```{r data values}
none <- 0.85
withdrawal <- 0.22
condom <- 0.18
pill <- 0.09
IUD <- 0.008
iud_progestin <- 0.002
female_ster <- 0.005
male_ster <- 0.0015
```

```{r functions}
FailRate <- function(years, method){
     pbinom(0, years, method, lower.tail = FALSE)
}

PlotFailRate <- function(years, method){
     rate <<- pbinom(0, 1:years, method, lower.tail = FALSE)
     p <- ggplot(as.data.frame(rate), aes(x = seq(1:length(rate)), y = rate))

     p +
     geom_line(aes(color = rate)) +
     scale_colour_gradient(high = "red")
}
```

```{r examples}
FailRate(4, condom)

FailRate(4, condom * pill)

PlotFailRate(4, condom)

PlotFailRate(4, condom * pill)
```

```{r data shape}
lifetime_failure <- tibble(years = 1:20,
                           none = FailRate(years, none),
                           withdrawal = FailRate(years, withdrawal),
                           condom = FailRate(years, condom),
                           pill = FailRate(years, pill),
                           IUD = FailRate(years, IUD))

failure_table <- pivot_longer(lifetime_failure, cols = -c(years), names_to = "method", values_to = "failure")

failure_table$method <- as.factor(failure_table$method)

failure_table$method <- fct_reorder(failure_table$method, failure_table$failure, .fun = mean, .desc = TRUE)
```

```{r plot}
failure_plot <- ggplot(failure_table) +
    aes(x = years, y = failure * 100, colour = method) +
    geom_smooth(size = 1, se = FALSE, method = "loess", formula = "y ~ x") +
    geom_hline(yintercept = 0) +
    scale_colour_viridis_d(option = "D", direction = 1) +
    scale_x_continuous(expand = expansion(0)) + 
    scale_y_continuous(expand = expansion(0)) +
    geom_hline(yintercept = 25, alpha = 0.25) +
    geom_hline(yintercept = 50, alpha = 0.25) +
    geom_hline(yintercept = 75, alpha = 0.25) +
    ggtitle("Probability of At Least One Contraceptive Failure by Method", subtitle = "With Typical Use") +
    xlab("Years of Use") +
    ylab("% Chance of at Least One Pregnancy") +
    theme_bw()

failure_plot
```

```{r print plot}
ragg::agg_png("failure_plot_16x9.png", width = 16, height = 9, units = "in", res = 300, scaling = 2.8)
failure_plot
dev.off()
```

## Quarto

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
